# EEB319 Lab 4: TA Copy

## Exercise 1

```{r}
r <- 0.1
K <- 1000
t <- seq(0, 100, 1)
N <- c(); N[1] <- 5

for (i in 2:length(t)){
  N[i] <- N[i-1]*exp(r*(1-(N[i-1]/K)))
}

plot(N, xlab = "Time", ylab = "Population Abundance (N)")
```

## Exercise 2

```{r}
r <- 0.1
K <- 1000
t <- seq(0, 100, 1)
N <- c(); N[1] <- 5
N2 <- c(); N2[1] <- 200

for (i in 2:length(t)){
  N[i] <- N[i-1]*exp(r*(1-(N[i-1]/K)))
  N2[i] <- N2[i-1]*exp(r*(1-(N2[i-1]/K)))
}

plot(N, xlab = "Time", ylab = "Population Abundance (N)", type = "l")
lines(N2, col = "red3")
legend("bottomright", legend = c("Initial Abundance = 5", "Initial Abundance = 200"),
       col = c("black", "red3"), lwd = 2)
```

How should we refer to this value?

*As the equilibrium abundance (or anything similar).*

What do you notice about this value?

*It's equal to K; the population plateaus at carrying capacity.*

## Exercise 3

```{r}
## Parameters
r <- 0.1
K <- 1000
sd <- 0.4

## Time
t <- seq(0, 100, 1)

## Empty vector for population abundance (N), and initial condition
N <- c(); N[1] <- 5

## Draw from normal distribution for error terms
set.seed(1) # So we all get the same answers!
epsilon <- rnorm(n = t - 1, mean = 0, sd = sd)

## Loop time!
for(i in 2:length(t)){
  N[i] <- N[i - 1]*exp(0.1*(1 - (N[i - 1]/K)) + epsilon[i - 1])
}

## Plotting
plot(N, type = "l", xlab = "Time", ylab = "Population Abundance (N)")
```

## Exercise 4

```{r}
## Parameters
r <- 0.1
K <- 1000
sd <- 0.4

## Time
t <- seq(0, 100, 1)

## Number of population simulations we want to do
pop_sims <- 5

## Draw from normal distribution for error terms
epsilon_df <- c()
for (i in 1:pop_sims){
  set.seed(i)
  epsilon <- rnorm(n = t - 1, mean = 0, sd = sd)
  epsilon_df <- cbind(epsilon_df, epsilon)
}

## Loop time! And this time, it's nested!
N_df <- c()
for (i in 1:pop_sims){
  N <- c(); N[1] <- 5
  for (j in 2:length(t)){
    N[j] <- N[j - 1]*exp(0.1*(1 - (N[j - 1]/K)) + epsilon_df[j - 1, i])
  }
  N_df <- cbind(N_df, N)
}

## Plotting
cols <- c("red3", "gold", "#009966", "#003399", "#CC0066")
plot(1, 1, pch = "", xlim = c(0, max(t)), ylim = c(0, max(N_df)), xlab = "Time", ylab = "Population Abundance (N)")
for (i in 1:ncol(N_df)){
  points(N_df[ ,i], col = cols[i], type = "l", lwd = 1)
}
```

## Exercise 5

```{r}
r <- c(0.5, 1.0, 1.5, 1.9, 2, 2.2, 2.5, 2.6, 2.7, 2.8, 2.9)
K <- 1000
t <- seq(0, 100, 1)

N_df <- c()
for (i in 1:length(r)){
  N <- c(); N[1] <- 5
  for (j in 2:length(t)){
    N[j] <- N[j-1]*exp(r[i]*(1-(N[j-1]/K)))
  }
  N_df <- cbind(N_df, N)
}

## Plotting
cols <- c("red3", "orange1", "gold", "#00CC99", "#009966", "#66CCFF", "#003399", "#330066", "#CC99FF", "#FF99FF", "#CC0066")
plot(1, 1, pch = "", xlim = c(0, 50), ylim = c(0, max(N_df)), xlab = "Time", ylab = "Population Abundance (N)")
for (i in 1:ncol(N_df)){
  points(N_df[ ,i], col = cols[i], type = "l", lwd = 1)
}

## That's a lot! Let's plot sequentially:
plot(N_df[ ,1], type = "l", col = cols[1], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 0.5", ylim = c(0, 2500))

plot(N_df[ ,2], type = "l", col = cols[2], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 1.0", ylim = c(0, 2500))

plot(N_df[ ,3], type = "l", col = cols[3], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 1.5", ylim = c(0, 2500))

plot(N_df[ ,4], type = "l", col = cols[4], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 1.9", ylim = c(0, 2500))

plot(N_df[ ,5], type = "l", col = cols[5], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 2.0", ylim = c(0, 2500))

plot(N_df[ ,6], type = "l", col = cols[6], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 2.2", ylim = c(0, 2500))

plot(N_df[ ,7], type = "l", col = cols[7], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 2.5", ylim = c(0, 2500))

plot(N_df[ ,8], type = "l", col = cols[8], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 2.6", ylim = c(0, 2500))

plot(N_df[ ,9], type = "l", col = cols[9], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 2.7", ylim = c(0, 2500))

plot(N_df[ ,10], type = "l", col = cols[10], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 2.8", ylim = c(0, 2500))

plot(N_df[ ,11], type = "l", col = cols[11], xlab = "Time", ylab = "Population Abundance (N)", main = "r = 2.9", ylim = c(0, 2500))
```

Describe how the dynamics change as you increase $r$.

*Dynamics start as pretty stable, then we see oscillatory instability, first as a stable limit cycle, then as bifurcations occur, we see the population oscillate between an increasing number of equilibria. They won't word it this way, but vibes count.*

## Exercise 6

```{r}
r <- c(2.9, 3.0)
K <- 1000
t <- seq(0, 100, 1)
N_init <- c(5, 5.1)

## r = 2.9
N_r1 <- c()
for (i in 1:length(N_init)){
  N <- c(); N[1] <- N_init[i]
  for (j in 2:length(t)){
    N[j] <- N[j-1]*exp(r[1]*(1-(N[j-1]/K)))
  }
  N_r1 <- cbind(N_r1, N)
}

## r = 3.0
N_r2 <- c()
for (i in 1:length(N_init)){
  N <- c(); N[1] <- N_init[i]
  for (j in 2:length(t)){
    N[j] <- N[j-1]*exp(r[2]*(1-(N[j-1]/K)))
  }
  N_r2 <- cbind(N_r2, N)
}

## Plotting
plot(N_r1[ ,1], col = "red3", type = "l", lwd = 2, lty = 1, xlab = "Time", ylab = "Population Abundance (N)", main = "r = 2.9", ylim = c(0, 2500))
lines(N_r1[ ,2], col = "#CC99FF", type = "l", lwd = 2, lty = 2)

plot(N_r2[ ,1], col = "red3", type = "l", lwd = 2, lty = 1, xlab = "Time", ylab = "Population Abundance (N)", main = "r = 3.0", ylim = c(0, 2500))
lines(N_r2[ ,2], col = "#CC99FF", type = "l", lwd = 2, lty = 2)
```

## Exercise 7

```{r}
## Parameters
f <- 12
a <- 5
s <- 0.85

## Time
t <- seq(1, 2000, 1)

## Empty vectors and initial conditions
J <- c(); J[1] <- 100
A <- c(); A[1] <- 120

## Simulation
for(i in 2:length(t)){
  J[i] <- A[i-1]*exp(f)
  A[i] <- J[i-1]*exp(-a*J[i-1]) + s*A[i-1]
}

## Ploting
plot(J[500:2000], type = "l", xlab = "Time", ylab = "Abundance", main = "Juvenile Abundance, Years 500-2000")
plot(A[500:2000], type = "l", xlab = "Time", ylab = "Abundance", main = "Adult Abundance, Years 500-2000")
```

What do you conclude from the plots?

*Something about looks pretty chaotic, but if we only saw the data, and not the underlying model, we might not be able to differentiate chaos from randomness.*

## Exercise 8

```{r}
plot(J[500:2000], A[500:2000], xlab = "Juvenile Abundance", ylab = "Adult Abundance", main = "State Space: Years 500-2000")
```

How do the individual time series look in comparison to the state space?

*Something about a clear pattern indicating that a deterministic process (and not a stochastic one), underlies these data.*

## Exercise 9

```{r}
J <- rnorm(n = 1500, mean = 100, sd = 10)
A <- rnorm(n = 1500, mean = 120, sd = 10)

plot(J, A, xlab = "Juvenile Abundance", ylab = "Adult Abundance", main = "Null Model State Space")
```

What does the state space look like? How does this compare with the state space from Exercise 10?

*Random! Rawr XD.*

## Exercise 10

Based on your null model, what conclusions can you draw from your realized juvenile-adult state space? Explain. Also, how do your conclusions relate to your expectations when you first looked at the individual time series of juveniles and adults?

*:)*
