# EEB319 Lab 3: Harvesting and Stability

## Exercise 1.1

```{r}
## Parameters
r <- 1 # Growth rate
K <- 1000 # Carrying capacity

## Biomass
N <- seq(1, 1000, 1)

## Empty vector
G <- c()

## Initial conditions
G[1] <- 0
```

```{r}
for (i in N){
  G[i] <- r*N[i]*(1 - (N[i]/K))
}
```

```{r}
plot(G, type = "l", lwd = 2,
     main = "Growth Function", xlab = "Population Abundance", ylab = "Growth Rate")
```

```{r}
r_1 <- 0.8
K <- 1000
N <- seq(1, 1000, 1)
G_1 <- c()
G_1[1] <- 0

for (i in N){
  G_1[i] <- r_1*N[i]*(1 - (N[i]/K))
}

plot(G_1, type = "l", lwd = 2,
     main = "Growth Function: Low Growth Rate", xlab = "Population Abundance", ylab = "Growth Rate")

r <- 1
K_2 <- 800
N <- seq(1, 1000, 1)
G_2 <- c()
G_2[1] <- 0

for (i in N){
  G_2[i] <- r*N[i]*(1 - (N[i]/K_2))
}

plot(G_2, type = "l", lwd = 2,
     main = "Growth Function: Low Carrying Capacity", xlab = "Population Abundance", ylab = "Growth Rate")
```

## Exercise 1.2

```{r}
e <- 0.3 # Slope
N <- seq(1, 1000, 1) # Biomass
H <- c(); H[1] <- 0 # Empty vector and first position
```

```{r}
for (i in N){
  H[i] <- e*N[i]
}
```

```{r}
plot(G, type = "l", lwd = 2,
     main = "Growth Function", xlab = "Population Abundance", ylab = "Growth Rate")
lines(H, lty = 2, lwd = 2, col = "red3")
legend("topleft", legend = c("Growth Function", "Harvest Function"),
       col = c("black", "red3"), lty = c(1,2), lwd = 2)
```

Equilibria:

1.  $N^*_1=0$
2.  $N^*_2=K(1-\frac{e}{r})$

## Exercise 1.3

```{r}
K <- 1000; r <- 1
e_seq <- seq(0, 1, 0.1)
N <- c()
for (i in 1:length(e_seq)){
  N[i] <- (K*(r - e_seq[i]))/r
}

df <- as.data.frame(cbind(e_seq, N))
plot(df$e_seq, df$N, type = "l", lwd = 2,
     main = "Population Abundance at Equilibrium", xlab = "Harvesting Rate", ylab = "Population Abundance")
abline(v = e, lwd = 2, lty = 2, col = "blue3")
```

## Exercise 2.1

```{r}
r <- 1
K <- 1000
H <- 150 # Constant harvesting

N <- seq(1, 1000, 1)

G <- c(); G[1]<-0

for (i in N){
  G[i] <- r*N[i]*(1 - (N[i]/K))
}

plot(G, type = "l", lwd = 2,
     main = "System Equilibria Under Constant Quota", xlab = "N", ylab = "Growth Function")
abline(h = H, lwd = 2, lty = 2, col = "red3")
legend("topright", legend = c("Growth Function", "Constant Quota"), lty = 1, lwd = 2, col = c("black", "red"))
```

Equilibria at intersections; unstable (left) and stable (right).

## Exercise 2.2

$H=\frac{rk}{4}$

## Exercise 2.3

When constant quota is 0, no equilibria. As the quota increases, the horizontal line from the plot above moves up the plot. As such, as the quota increases, the equilibria points move closer together, until you see a single equilibrium point at $\frac{k}{2}$. As a plot, this would look like two lines, one starting at $N=0$, and the second starting at $N=K$, with positive and negative slopes, respectively, moving linearly until they meet at $N=\frac{k}{2}$.

## Exercise 3.1

Blah, blah, blah.

## Exercise 3.2

Blah, blah, blah.
