# EEB319 Lab 5: Stage-Structured Population Dynamics

## Part 1

### Exercise 1.1

Using the life history graph, write down two equations for $J_{t+1}$ and $A_{t+1}$.

$$
\begin{align}
J_{t+1}&=A_tS_Af \\
A_{t+1}&=A_tS_A+J_tS_J
\end{align}
$$

### Exercise 1.2

Write down the corresponding population projection matrix (PPM).

|                        | $J_{t+1}$ or $J_t$ | $A_{t+1}$ or $A_t$ |
|------------------------|--------------------|--------------------|
| $J_t$ **or** $J_{t+1}$ | $0$                | $S_Af$             |
| $A_t$ **or** $A_{t+1}$ | $S_J$              | $S_A$              |

### Exercise 1.3

List three assumptions of the stage-structured model.

*Blah, blah, blah.*

### Exercise 1.4

```{r}
## Parameters
S_J <- 0.1
S_A <- 0.8
f <- 3.5

## Time
t <- seq(0, 40, 1)

## Empty vectors and initial conditions
J <- c(); J[1] <- 10
A <- c(); A[1] <- 0

## For loop
for (i in 2:length(t)){
  J[i] <- A[i-1]*S_A*f
  A[i] <- A[i-1]*S_A + J[i-1]*S_J
}

## Plotting
plot(J, type = "l", col = "orchid1", lwd = 2, xlab = "Time", ylab = "Population Abundance")
lines(A, col = "red3", lwd = 2)
legend("topleft", legend = c("Juveniles", "Adults"), col = c("orchid1", "red3"), lwd = 2)
```

### Exercise 1.5

```{r}
## Matrix to make this easier to conceptualise
popn_matrix <- cbind(t, J, A)
head(popn_matrix)

## Empty vectors
J_lambdas <- c(); A_lambdas <- c()

## For loop
for (i in 2:nrow(popn_matrix)){
  J_lambdas[i-1] <- popn_matrix[i, 2]/popn_matrix[i-1, 2]
  A_lambdas[i-1] <- popn_matrix[i, 3]/popn_matrix[i-1, 3]
}

## Growth rate at requested time points
J_lambdas[20]; J_lambdas[30]; J_lambdas[40]
A_lambdas[20]; A_lambdas[30]; A_lambdas[40]
```

What is the value of $\lambda$ for adults and juveniles at these time points? Do they differ?

*Nope.*

### Exercise 1.6

```{r}
total_popn <- rowSums(popn_matrix[ ,2:3])
head(total_popn)

J_prop <- popn_matrix[ ,2]/total_popn
head(J_prop)

J_prop[20]; J_prop[30]; J_prop[40]
```

Do they differ through time?

*Nope.*

### Exercise 1.7

```{r}
plot(J, A, type = "b", xlab = "Juvenile Abundance", ylab = "Adult Abundance", main = "Phase Plane")
```

What pattern do you see and how does it relate to the eigenvalue and eigenvector?

*Blah, blah, blah.*

## Part 2

### Exercise 2.1

$$
\begin{align}
n_{1, t+1} &= F_2n_{2,t} + F_3n_{3,t} \\
n_{2, t+1} &= p_{22}n_{2,t} + p_{21}n_{1,t} \\
n_{3, t+1} &= p_{33}n_{3,t} + p_{32}n_{2,t} + p34n_{4,t} \\
n_{4, t+1} &= p_{42}n_{2,t} + p_{43}n_{3,t} 
\end{align}
$$

### Exercise 2.2

```{r}
## Parameters
p21 <- 0.92; p22 <- 0.86
p32 <- 0.08; p33 <- 0.80; p34 <- 0.75
p42 <- 0.02; p43 <- 0.19

## Fecundity equations
F2 <- 0.5*p42*sqrt(p34); F3 <- 0.5*p43*sqrt(p34)

## Time
t <- seq(0, 100, 1)

## Empty vectors and initial conditions
n1 <- c(); n1[1] <- 200
n2 <- c(); n2[1] <- 200
n3 <- c(); n3[1] <- 200
n4 <- c(); n4[1] <- 200

## Model simulation via for loop
for (i in 2:length(t)){
  n1[i] <- F2*n2[i-1] + F3*n3[i-1]
  n2[i] <- p22*n2[i-1] + p21*n1[i-1]
  n3[i] <- p33*n3[i-1] + p32*n2[i-1] + p34*n4[i-1]
  n4[i] <- p42*n2[i-1] + p43*n3[i-1]
}

popn_matrix <- cbind(t, n1, n2, n3, n4)
head(popn_matrix)

## Plotting
plot(n1, type = "l", col = "lightpink", lwd = 2, xlab = "Time", ylab = "Population Abundance", ylim = c(0, max(popn_matrix)))
lines(n2, col = "#00CC99", lwd = 2)
lines(n3, col = "lightblue", lwd = 2)
lines(n4, col = "#CC99FF", lwd = 2)
legend("topright", legend = c("Calves", "Immature females", "Mature females", "Mature females w/ calves"), col = c("lightpink", "#00CC99", "lightblue", "#CC99FF"), lwd = 2)
```

```{r}
## Empty vectors
n1_lambdas <- c(); n2_lambdas <- c(); n3_lambdas <- c(); n4_lambdas <- c()

## For loop
for (i in 2:nrow(popn_matrix)){
  n1_lambdas[i-1] <- popn_matrix[i, 2]/popn_matrix[i-1, 2]
  n2_lambdas[i-1] <- popn_matrix[i, 3]/popn_matrix[i-1, 3]
  n3_lambdas[i-1] <- popn_matrix[i, 4]/popn_matrix[i-1, 4]
  n4_lambdas[i-1] <- popn_matrix[i, 5]/popn_matrix[i-1, 5]
}

plot(n1_lambdas, type = "l", col = "lightpink", lwd = 2, lty = 1, xlab = "Time", ylab = "Lambda")
lines(n2_lambdas, col = "#00CC99", lwd = 2, lty = 2)
lines(n3_lambdas, col = "lightblue", lwd = 2, lty = 3)
lines(n4_lambdas, col = "#CC99FF", lwd = 2, lty = 4)
legend("topright", legend = c("Calves", "Immature females", "Mature females", "Mature females w/ calves"), col = c("lightpink", "#00CC99", "lightblue", "#CC99FF"), lwd = 2)

n1_lambdas[length(n1_lambdas)]
n2_lambdas[length(n2_lambdas)]
n3_lambdas[length(n3_lambdas)]
n4_lambdas[length(n4_lambdas)]
```

*Geometric population growth rate = 0.998134*

```{r}
total_popn <- rowSums(popn_matrix[ ,2:5])
head(total_popn)

n1_prop <- popn_matrix[ ,2]/total_popn
n2_prop <- popn_matrix[ ,3]/total_popn
n3_prop <- popn_matrix[ ,4]/total_popn
n4_prop <- popn_matrix[ ,5]/total_popn

plot(n1_prop, type = "l", col = "lightpink", lwd = 2, lty = 1, xlab = "Time", ylab = "Lambda", ylim = c(0, 1))
lines(n2_prop, col = "#00CC99", lwd = 2, lty = 2)
lines(n3_prop, col = "lightblue", lwd = 2, lty = 3)
lines(n4_prop, col = "#CC99FF", lwd = 2, lty = 4)
legend("topright", legend = c("Calves", "Immature females", "Mature females", "Mature females w/ calves"), col = c("lightpink", "#00CC99", "lightblue", "#CC99FF"), lwd = 2)

n1_prop[length(n1_prop)]
n2_prop[length(n2_prop)]
n3_prop[length(n3_prop)]
n4_prop[length(n4_prop)]
```

*Steady-stage distribution:*

$n_1$ = 0.04672768

$n_2$ = 0.3112158

$n_3$ = 0.5341436

$n_4$ = 0.107913

### Exercise 2.3

```{r}
## Value of p34
p34 <- 0.77

## Fecundity equations
F2 <- 0.5*p42*sqrt(p34); F3 <- 0.5*p43*sqrt(p34)

## Empty vectors and initial conditions
n1 <- c(); n1[1] <- 200
n2 <- c(); n2[1] <- 200
n3 <- c(); n3[1] <- 200
n4 <- c(); n4[1] <- 200

## Model simulation via for loop
for (i in 2:length(t)){
  n1[i] <- F2*n2[i-1] + F3*n3[i-1]
  n2[i] <- p22*n2[i-1] + p21*n1[i-1]
  n3[i] <- p33*n3[i-1] + p32*n2[i-1] + p34*n4[i-1]
  n4[i] <- p42*n2[i-1] + p43*n3[i-1]
}

popn_matrix <- cbind(t, n1, n2, n3, n4)

## Finding geometric population growth rate
n1_lambdas <- c(); n2_lambdas <- c(); n3_lambdas <- c(); n4_lambdas <- c()

for (i in 2:nrow(popn_matrix)){
  n1_lambdas[i-1] <- popn_matrix[i, 2]/popn_matrix[i-1, 2]
  n2_lambdas[i-1] <- popn_matrix[i, 3]/popn_matrix[i-1, 3]
  n3_lambdas[i-1] <- popn_matrix[i, 4]/popn_matrix[i-1, 4]
  n4_lambdas[i-1] <- popn_matrix[i, 5]/popn_matrix[i-1, 5]
}

n1_lambdas[length(n1_lambdas)]; n2_lambdas[length(n2_lambdas)]; n3_lambdas[length(n3_lambdas)]; n4_lambdas[length(n4_lambdas)]
```

$p_{34}$ *must be greater than or equal to 0.77 (using same degree of precision as given in question).*

```{r}
n4_SSD <- n4_prop[length(n4_prop)]; 77*n4_SSD
```

*There are currently 8 (8.309298) mothers with calves.*

*Number of mothers that need to be saved to push population growth rate over 1: 1 (I don't have a great way to do this actually because I'm confused by the question, but I think this is the right answer).*
