# EEB319 Lab 7: Infectious Disease Modeling

## Exercise 1

```{r}
## Parameters
b <- 0.05
k <- 50
dt <- 0.001

## Time
t <- seq(0, 100, 1)

## Empty vectors and initial conditions
S <- c(); S[1] <- 10000
I <- c(); I[1] <- 1
R <- c(); R[1] <- 0

## Model simulation via for loop
for(i in 2:length(t)){
  S[i] <- -b*S[i-1]*I[i-1]*dt + S[i-1]
  I[i] <- (b*S[i-1]*I[i-1]-k*I[i-1])*dt + I[i-1]
  R[i] <- k*I[i-1]*dt + R[i-1]
}

## Plotting
plot(S, xlab = "Time", ylab = "Population Abundance", type = "l")
lines(I, lty = 2) 
lines(R, lty = 3)
legend("topright", legend = c("Susceptible", "Infected", "Removed"), lty = c(1, 2, 3))
```

Describe the trends for each population category within the model. Are the trends as you expected? Why or why not? Does the number of infected individuals peak? If so, when?

*Blah, blah, blah.*

## Exercise 2

```{r}
## Calculating R0 for above simulation
R0 <- b*S[1]/k; R0

## Calculating Host Density Threshold (NC) for simulation above
NC <- k/b; NC

#### New simulations! ####
## Parameters
b <- 0.05
k <- 50
dt <- 0.001

## Time
t <- seq(0, 100, 1)

## Model simulation via for loop
S_concat <- c(); I_concat <- c(); R_concat <- c()
S_inits <- c(5000, 4000, 2000, 1001, 1000, 500)
for (i in 1:length(S_inits)){
  S <- c(); S[1] <- S_inits[i]
  I <- c(); I[1] <- 1
  R <- c(); R[1] <- 0
  for (j in 2:length(t)){
  S[j] <- -b*S[j-1]*I[j-1]*dt + S[j-1]
  I[j] <- (b*S[j-1]*I[j-1]-k*I[j-1])*dt + I[j-1]
  R[j] <- k*I[j-1]*dt + R[j-1]
  }
  S_concat <- cbind(S_concat, S)
  I_concat <- cbind(I_concat, I)
  R_concat <- cbind(R_concat, R)
}

## Calculating R0
R0s <- c()
for (i in 1:ncol(S_concat)){
  R0s[i] <- b*S_concat[1, i]/k
}
R0s

## Plotting
plot(S_concat[, 1], xlab = "Time", ylab = "Population Abundance", main = "Initial # of Susceptible Individuals = 5000", type = "l", ylim = c(0, 5000))
lines(I_concat[, 1], lty = 2) 
lines(R_concat[, 1], lty = 3)
legend("topright", legend = c("Susceptible", "Infected", "Removed"), lty = c(1, 2, 3))

plot(S_concat[, 2], xlab = "Time", ylab = "Population Abundance", main = "Initial # of Susceptible Individuals = 4000", type = "l", ylim = c(0, 5000))
lines(I_concat[, 2], lty = 2) 
lines(R_concat[, 2], lty = 3)
legend("topright", legend = c("Susceptible", "Infected", "Removed"), lty = c(1, 2, 3))

plot(S_concat[, 3], xlab = "Time", ylab = "Population Abundance", main = "Initial # of Susceptible Individuals = 2000", type = "l", ylim = c(0, 5000))
lines(I_concat[, 3], lty = 2) 
lines(R_concat[, 3], lty = 3)
legend("topright", legend = c("Susceptible", "Infected", "Removed"), lty = c(1, 2, 3))

plot(S_concat[, 4], xlab = "Time", ylab = "Population Abundance", main = "Initial # of Susceptible Individuals = 1001", type = "l", ylim = c(0, 5000))
lines(I_concat[, 4], lty = 2) 
lines(R_concat[, 4], lty = 3)
legend("topright", legend = c("Susceptible", "Infected", "Removed"), lty = c(1, 2, 3))

plot(S_concat[, 5], xlab = "Time", ylab = "Population Abundance", main = "Initial # of Susceptible Individuals = 1000", type = "l", ylim = c(0, 5000))
lines(I_concat[, 5], lty = 2) 
lines(R_concat[, 5], lty = 3)
legend("topright", legend = c("", "Infected", "Removed"), lty = c(1, 2, 3))

plot(S_concat[, 6], xlab = "Time", ylab = "Population ", main = "Initial # of Susceptible Individuals = 500", type = "l", ylim = c(0, 5000))
lines(I_concat[, 6], lty = 2) 
lines(R_concat[, 6], lty = 3)
legend("topright", legend = c("Susceptible", "Infected", "Removed"), lty = c(1, 2, 3))
```

How do the patterns change as you alter $N$? Although probably not visually apparent, how do the plots that begin with 1001 and 1000 susceptible individuals differ?

*Blah, blah, blah.*

## Exercise 3

For $\beta$:

$$
\begin{align}
R_0&=\frac{\beta S_0}{k} \\
1&=\frac{\beta 10,000}{50} \\
50&=\beta 10,000 \\
\beta&=0.005
\end{align}
$$

For $k$:

$$
\begin{align}
R_0&=\frac{\beta S_0}{k} \\
1&=\frac{0.05*10,000}{k} \\
k&=0.05*10,000 \\
k&=500
\end{align}
$$

```{r}
#### Original simulation ####
## Parameters
b <- 0.05
k <- 50
dt <- 0.001

## Time
t <- seq(0, 100, 1)

## Empty vectors and initial conditions
S <- c(); S[1] <- 10000
I <- c(); I[1] <- 1
R <- c(); R[1] <- 0

## Model simulation via for loop
for(i in 2:length(t)){
  S[i] <- -b*S[i-1]*I[i-1]*dt + S[i-1]
  I[i] <- (b*S[i-1]*I[i-1]-k*I[i-1])*dt + I[i-1]
  R[i] <- k*I[i-1]*dt + R[i-1]
}

## R0
R0 <- b*S[1]/k; R0

## NC
NC <- k/b; NC

## Plotting
plot(S, xlab = "Time", ylab = "Population Abundance", main = expression(paste("β = 0.05; k = 50; ", R[0], " = 10")), type = "l")
lines(I, lty = 2) 
lines(R, lty = 3)
legend("topright", legend = c("Susceptible", "Infected", "Removed"), lty = c(1, 2, 3))

#### Decreased beta simulation ####
b <- 0.005
k <- 50
dt <- 0.001

t <- seq(0, 100, 1)

S <- c(); S[1] <- 10000
I <- c(); I[1] <- 1
R <- c(); R[1] <- 0

for(i in 2:length(t)){
  S[i] <- -b*S[i-1]*I[i-1]*dt + S[i-1]
  I[i] <- (b*S[i-1]*I[i-1]-k*I[i-1])*dt + I[i-1]
  R[i] <- k*I[i-1]*dt + R[i-1]
}

R0 <- b*S[1]/k; R0
NC <- k/b; NC

plot(S, xlab = "Time", ylab = "Population Abundance", main = expression(paste("β = 0.005; k = 50; ", R[0], " = 1")), type = "l")
lines(I, lty = 2) 
lines(R, lty = 3)
legend("topright", legend = c("Susceptible", "Infected", "Removed"), lty = c(1, 2, 3))

#### Increased k simulation ####
b <- 0.05
k <- 500
dt <- 0.001

t <- seq(0, 100, 1)

S <- c(); S[1] <- 10000
I <- c(); I[1] <- 1
R <- c(); R[1] <- 0

for(i in 2:length(t)){
  S[i] <- -b*S[i-1]*I[i-1]*dt + S[i-1]
  I[i] <- (b*S[i-1]*I[i-1]-k*I[i-1])*dt + I[i-1]
  R[i] <- k*I[i-1]*dt + R[i-1]
}

R0 <- b*S[1]/k; R0
NC <- k/b; NC

plot(S, xlab = "Time", ylab = "Population Abundance", main = expression(paste("β = 0.05; k = 500; ", R[0], " = 1")), type = "l")
lines(I, lty = 2) 
lines(R, lty = 3)
legend("topright", legend = c("Susceptible", "Infected", "Removed"), lty = c(1, 2, 3))
```

Describe your own individual strategy where you alter all three parameters to eradicate the disease in the population. How could these parameters be met?

*Blah, blah, blah.*
